\textbf{Effective application of machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-09-10 submission/manuscript.Rmd
add links

**16S rRNA gene sequencing data.** Stool samples provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances. The sequence data used in our analyses were originally generated by Baxter et al. (available through NCBI Sequence Read Archive [SRP062005], 2015). The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm ([@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; GitHub link to table). These OTU abundances were the features we used to predict colorectal health of the patients. There were 6920 OTUs. OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1 and lowest would be 0.

**Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to calculate decision values instead of predicted probabilities for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel. These changes were necessary to calculate AUROC values for SVMs. The code for these changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/.

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. The range of the grid changes with different ML tasks and ML models. A full grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the _cost_ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned _sigma_ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the _depth of the tree_ where deeper the tree, the more splits it has. For random forest, we tuned the _number of features_ to consider when looking for the best tree split. For XGBoost, we tuned for _learning rate_ and the _fraction of samples_ to be used for fitting the individual base learners. There are newly developed tools such as Hyperband that help researcher with hyperparameter selection [@li_hyperband:_2016] that can be incorporated to microbiome studies.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.

>>>>

**16S rRNA gene sequencing data.** Stool samples provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances. The sequence data used in our analyses were originally generated by Baxter et al. (available through NCBI Sequence Read Archive [SRP062005], 2015). The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}). These OTU abundances were the features we used to predict colorectal health of the patients. There were 6920 OTUs. OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1 and lowest would be 0.

**Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to calculate decision values instead of predicted probabilities for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel. These changes were necessary to calculate AUROC values for SVMs. The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. The range of the grid changes with different ML tasks and ML models. A full grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners. There are newly developed tools such as Hyperband that help researcher with hyperparameter selection [@li_hyperband:_2016] that can be incorporated to microbiome studies.

@@

**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.

### 2019-09-10 submission/manuscript.tex
add links

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective model selection and implementation when applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective model selection and implementation when applying
machine learning to bacterial 16S rRNA gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients was used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (32), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm ((37--39);
GitHub link to table). These OTU abundances were the features we used to
predict colorectal health of the patients. There were 6920 OTUs. OTU
abundances were subsampled to the size of the smallest sample and
normalized across samples such that the highest abundance of each OTU
would be 1 and lowest would be 0.

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values instead of predicted probabilities for models
generated using L2-regularized SVM with linear kernel and L1-regularized
SVM with linear kernel. These changes were necessary to calculate AUROC
values for SVMs. The code for these changes are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/}.

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. The
range of the grid changes with different ML tasks and ML models. A full
grid search was needed to avoid large variation in prediction
performance. For L2-regularized logistic regression, L1 and
L2-regularized SVM with linear and radial basis function kernels, we
tuned the \emph{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \emph{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \emph{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\emph{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \emph{learning rate} and the
\emph{fraction of samples} to be used for fitting the individual base
learners. There are newly developed tools such as Hyperband that help
researcher with hyperparameter selection (40) that can be incorporated
to microbiome studies.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{\textbf{Effective application of machine learning to bacterial 16S rRNA
gene sequencing data}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients was used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (32), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (37--39);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1 and
lowest would be 0.

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values instead of predicted probabilities for models
generated using L2-regularized SVM with linear kernel and L1-regularized
SVM with linear kernel. These changes were necessary to calculate AUROC
values for SVMs. The code for these changes on L2-regularized SVM with
linear kernel and L1-regularized SVM with linear kernel are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
respectively.

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. The
range of the grid changes with different ML tasks and ML models. A full
grid search was needed to avoid large variation in prediction
performance. For L2-regularized logistic regression, L1 and
L2-regularized SVM with linear and radial basis function kernels, we
tuned the \textbf{cost} hyperparameter which determines the
regularization strength where smaller values specify stronger
regularization. For SVM with radial basis function kernel we also tuned
\textbf{sigma} hyperparameter which determines the reach of a single
training instance where for a high value of sigma, the SVM decision
boundary will be dependent on the points that are closest to the
decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For XGBoost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. There are newly developed
tools such as Hyperband that help researcher with hyperparameter
selection (40) that can be incorporated to microbiome studies.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
